---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

The [`CohortSurvival`](/docs/data_analysis/package_reference/cohortsurvival) and [`CohortConstructor`](/docs/data_analysis/package_reference/cohortconstructor) packages provide tools for executing comparative cohort time-to-event analyses within the OMOP Common Data Model.

## How Comparative Cohort Survival Works

- **Active Comparator Design**: Define target and comparator treatment cohorts using standard OMOP concepts with `CohortConstructor`.
- **New-User Requirement**: Restrict entries to incident users with a defined lookback period (`requirePriorObservation`).
- **Survival Estimation**: Compute non-parametric Kaplan-Meier survival curves and cumulative event probabilities using `estimateSingleEventSurvival()`.
- **Stratification & Visualisation**: Generate publication tables and faceted survival plots using `tableSurvival()` and `plotSurvival()`.

## Step 1: Setup & Connect to GiBleed

Load the necessary libraries and connect to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(CohortSurvival)
library(visOmopResults)
library(dplyr)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Build Target, Comparator, and Outcome Cohorts

We define new users of **Celecoxib** (`concept_id = 1118084`) as the target exposure, **Diclofenac** (`concept_id = 1124300`) as the active comparator exposure, and **Gastrointestinal Hemorrhage** (`concept_id = 192671`) as the safety outcome:

```{r build-cohorts, message=FALSE, warning=FALSE}
# 1. Instantiate Exposure Cohorts (Target & Active Comparator)
cdm$exposure <- conceptCohort(
  cdm = cdm,
  conceptSet = list(
    celecoxib = 1118084L,
    diclofenac = 1124300L
  ),
  name = "exposure"
) |>
  requireIsFirstEntry() |>
  requirePriorObservation(minPriorObservation = 365)

# 2. Instantiate Outcome Cohort (Gastrointestinal Hemorrhage)
cdm$outcome <- conceptCohort(
  cdm = cdm,
  conceptSet = list(gi_bleed = 192671L),
  name = "outcome"
)
```

## Step 3: Estimate Time-to-Event Survival

We estimate single-event survival probability comparing the risk of gastrointestinal hemorrhage between the two treatments:

```{r estimate-survival, message=FALSE, warning=FALSE}
# Estimate Kaplan-Meier survival curves
surv_estimates <- estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "exposure",
  outcomeCohortTable = "outcome"
)
```

## Step 4: Display Survival Tables and Event Summaries

We generate structured summaries of survival estimates and observed outcome event counts across follow-up time:

```{r survival-tables, message=FALSE, warning=FALSE}
# Summary table of survival probabilities
tableSurvival(surv_estimates)

# Table of outcome events and person-time
tableSurvivalEvents(surv_estimates)
```

## Step 5: Plot Kaplan-Meier Curves

We visualize the comparative Kaplan-Meier curves for Celecoxib versus Diclofenac users:

```{r plot-survival, message=FALSE, warning=FALSE}
# Plot comparative Kaplan-Meier survival curves
plotSurvival(surv_estimates, colour = "target_cohort")
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
