---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

The [`DrugUtilisation`](/docs/data_analysis/package_reference/drugutilisation) package provides standardized workflows to characterize real-world prescription patterns, treatment duration, dosage, indications, and longitudinal medication adherence from OMOP CDM databases.

## How DrugUtilisation works

- **New-User Cohorts**: Generate drug exposure cohorts with customizable washout and prior observation criteria using `CohortConstructor`.
- **Utilisation Metrics**: Calculate cumulative exposure duration, number of treatment eras, and quantity using `summariseDrugUtilisation()`.
- **Indication Analysis**: Identify recorded clinical diagnoses prior to therapy initiation using `summariseIndication()`.
- **Adherence & Persistence**: Measure treatment continuity via Proportion of Patients Covered (PPC) curves using `summariseProportionOfPatientsCovered()`.

## Step 1: Setup & Connect to GiBleed

First, load the required libraries and establish a connection to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(DrugUtilisation)
library(visOmopResults)
library(dplyr)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Define New-User Drug Cohort

We define a cohort of new users of **Celecoxib** (`concept_id = 1118084`) with at least 365 days of prior observation:

```{r define-drug-cohort, message=FALSE, warning=FALSE}
# Create new-user Celecoxib cohort with 365 days prior observation
cdm$celecoxib <- conceptCohort(
  cdm = cdm,
  conceptSet = list(celecoxib = 1118084L),
  name = "celecoxib"
) |>
  requireIsFirstEntry() |>
  requirePriorObservation(minPriorObservation = 365)
```

## Step 3: Summarise Drug Utilisation Metrics

We calculate key utilisation metrics including duration of exposure, number of eras, and cumulative quantity:

```{r summarise-drug-utilisation, message=FALSE, warning=FALSE}
# Summarise drug utilisation metrics
dus_summary <- cdm$celecoxib |>
  summariseDrugUtilisation(
    ingredientConceptId = 1118084L
  )

# Display publication table
tableDrugUtilisation(dus_summary)
```

## Step 4: Indication Profiling

We profile baseline clinical indications (such as acute sinusitis, UTI, and asthma) recorded in the 365-day and 30-day windows prior to drug initiation:

```{r summarise-indications, message=FALSE, warning=FALSE}
# Define potential indication cohorts
cdm$indications <- conceptCohort(
  cdm = cdm,
  conceptSet = list(
    sinusitis = 4283893L,
    uti = 4116491L,
    asthma = 4051466L
  ),
  name = "indications"
)

# Summarise indications prior to treatment start
ind_summary <- cdm$celecoxib |>
  summariseIndication(
    indicationCohortName = "indications",
    indicationWindow = list(c(-365, 0), c(-30, 0))
  )

# Display indication table
tableIndication(ind_summary)
```

## Step 5: Proportion of Patients Covered (Adherence)

We compute and visualize the Proportion of Patients Covered (PPC) over a 365-day post-index horizon to examine longitudinal treatment persistence:

```{r plot-ppc, message=FALSE, warning=FALSE}
# Calculate PPC over 365 days of follow-up
ppc <- cdm$celecoxib |>
  summariseProportionOfPatientsCovered(
    followUpDays = 365
  )

# Plot longitudinal patient coverage curve
plotProportionOfPatientsCovered(ppc)
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
