---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

The [`HERMES`](/docs/data_analysis/package_reference/hermes) package provides an end-to-end analytical framework for Health Economics and Outcomes Research (HEOR), Healthcare Resource Utilisation (HCRU) evaluation, and Health Technology Assessment (HTA) decision-analytic modeling directly from OMOP CDM databases.

## How the 6-Stage HEOR Pipeline Works

1. **Study Initialisation & Baseline Characterisation**: Defines comparative intervention arms and extracts demographic profiles.
2. **HCRU Extraction**: Measures encounter rates and lengths of stay across inpatient admissions, emergency visits, outpatient consultations, pharmacotherapy, and procedures.
3. **Causal Propensity Score Adjustment**: Adjusts for confounding by indication via regularized logistic regression and greedy caliper matching.
4. **Trajectory Compilation**: Translates longitudinal patient clinical journeys into discrete Markov state-transition matrices.
5. **Economic Simulation (Markov PSA)**: Executes probabilistic sensitivity analysis sampling Gamma cost distributions and Beta health state utility weights over a defined time horizon.
6. **Decision Analysis (CEA)**: Estimates the Incremental Cost-Effectiveness Ratio (ICER), Net Monetary Benefit (NMB), Cost-Effectiveness Acceptability Curves (CEAC), and Cost-Effectiveness Planes.

## Step 1: Setup & Connect to GiBleed

Load the necessary libraries and establish a connection to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(HERMES)
library(CDMConnector)
library(CohortConstructor)
library(dplyr)
library(gt)
library(ggplot2)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Define Target, Comparator, and Safety Outcome Cohorts

We instantiate new users of **Celecoxib** (`concept_id = 1118084`) as the target intervention, new users of **Diclofenac** (`concept_id = 1124300`) as the active comparator, and incident **Gastrointestinal Hemorrhage** (`concept_id = 192671`) as the primary health economic outcome:

```{r define-cohorts, message=FALSE, warning=FALSE}
# Target Cohort: Celecoxib new users
cdm$target_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = list(celecoxib = 1118084L),
  name = "target_cohort"
) |>
  requireIsFirstEntry()

# Comparator Cohort: Diclofenac new users
cdm$comparator_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = list(diclofenac = 1124300L),
  name = "comparator_cohort"
) |>
  requireIsFirstEntry()

# Outcome Cohort: Gastrointestinal Hemorrhage
cdm$outcome_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = list(gi_bleed = 192671L),
  name = "outcome_cohort"
)
```

## Step 3: Baseline Characterisation & HCRU Extraction

We initialize the study, characterize baseline demographics, and extract longitudinal healthcare resource utilization across baseline ($[-365, -1]$ days) and follow-up ($[0, 365]$ days) windows:

```{r extract-hcru, message=FALSE, warning=FALSE}
# Initialize HEOR study and extract HCRU
study <- init(
  cdm = cdm,
  target_cohort = "target_cohort",
  comparator_cohort = "comparator_cohort",
  outcome_cohort = "outcome_cohort"
) |>
  summarise_baseline() |>
  extract_hcru(
    baseline_window = c(-365, -1),
    followup_window = c(0, 365)
  )

# Summarise per-patient resource utilization rates
hcru_summary <- study$hcru$patient_summary |>
  group_by(window) |>
  summarise(
    Mean_Inpatient_Admissions = round(mean(inpatient_admissions), 2),
    Mean_Inpatient_LOS_Days = round(mean(inpatient_los_days), 2),
    Mean_Prescription_Fills = round(mean(prescription_fills), 2),
    Mean_Procedure_Count = round(mean(procedure_count), 2),
    .groups = "drop"
  )

hcru_summary |>
  gt() |>
  tab_header(
    title = "Healthcare Resource Utilisation (HCRU) Summary",
    subtitle = "Resource Consumption Across Baseline vs Follow-up Windows"
  ) |>
  cols_label(
    window = "Observation Window",
    Mean_Inpatient_Admissions = "Inpatient Admissions (Mean)",
    Mean_Inpatient_LOS_Days = "Length of Stay Days (Mean)",
    Mean_Prescription_Fills = "Prescription Fills (Mean)",
    Mean_Procedure_Count = "Procedures & Tests (Mean)"
  )
```

## Step 4: Causal Propensity Score Matching & Trajectory Compilation

We fit a regularized logistic regression propensity score model, perform 1:1 nearest-neighbor matching, and compile Markov state-transition probability matrices:

```{r ps-and-trajectories, message=FALSE, warning=FALSE}
# Causal propensity score adjustment and discrete trajectory modeling
study <- study |>
  fit_ps() |>
  adjust_ps(caliper = 0.2) |>
  compile_trajectories()

# Assign parametric state-specific unit costs (Baseline maintenance vs GI Bleed event)
study$costs <- data.frame(
  health_state = c("State_Baseline", "State_Outcome"),
  mean_cost = c(450, 4800),
  se_cost = c(40, 350)
)
```

## Step 5: Economic Simulation (Markov PSA) & Cost-Effectiveness Analysis

We execute a 10-year Monte Carlo Probabilistic Sensitivity Analysis (PSA) with 3% annual discounting, followed by decision analysis:

```{r simulate-and-cea, message=FALSE, warning=FALSE}
# Execute Probabilistic Sensitivity Analysis simulation
sim <- simulate_economics(
  traj_obj = study,
  time_horizon = 10,
  discount_rate = 0.03,
  n_samples = 250
)

# Run Cost-Effectiveness Decision Analysis
cea <- run_cea(sim)
```

## Step 6: HTA Decision Plots (CEAC & Cost-Effectiveness Plane)

We generate the Cost-Effectiveness Acceptability Curve (CEAC) showing the probability of cost-effectiveness across Willingness-to-Pay thresholds, and the Cost-Effectiveness Plane:

```{r plot-ceac, message=FALSE, warning=FALSE}
# Plot Cost-Effectiveness Acceptability Curve (CEAC)
plot_ceac(cea)
```

```{r plot-plane, message=FALSE, warning=FALSE}
# Plot Cost-Effectiveness Plane
plot_plane(cea)
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
