---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

Impact Evaluation Studies utilize quasi-experimental designs—such as **Interrupted Time Series (ITS)** and **Difference-in-Differences (DiD)**—to evaluate the causal impact of population-level health policies, regulatory safety warnings, or clinical guideline shifts on longitudinal disease rates or medication usage.

## How Interrupted Time Series Works

- **Longitudinal Rate Series**: Tracking population-level incidence or prevalence rates over time across pre- and post-intervention periods.
- **Intervention Threshold**: A known calendar cutoff date (e.g. year of a regulatory risk minimization measure).
- **Segmented Regression**: Modeling the time series piecewise to test for:
  - **Step Change ($\beta_2$)**: Immediate level jump or drop in outcome rate following the intervention.
  - **Slope Change ($\beta_3$)**: Change in the long-term rate trajectory or gradient post-intervention.

## Step 1: Setup & Connect to GiBleed

Load the required libraries and connect to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(IncidencePrevalence)
library(visOmopResults)
library(dplyr)
library(ggplot2)
library(gt)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Define Outcome and Longitudinal Denominator

We track annual incidence rates of **Gastrointestinal Hemorrhage** (`concept_id = 192671`) across the source population observed from 1980 through 2019:

```{r define-cohorts, message=FALSE, warning=FALSE}
# 1. Instantiate Outcome Cohort
cdm$outcome <- conceptCohort(
  cdm = cdm,
  conceptSet = list(gi_bleed = 192671L),
  name = "outcome"
)

# 2. Instantiate Longitudinal Study Denominator
cdm <- generateDenominatorCohortSet(
  cdm = cdm,
  name = "denominator",
  cohortDateRange = as.Date(c("1980-01-01", "2019-01-01"))
)
```

## Step 3: Estimate Longitudinal Incidence Rates

We calculate annual incidence rates per 100,000 person-years:

```{r estimate-rates, message=FALSE, warning=FALSE}
# Estimate annual incidence rates
inc <- estimateIncidence(
  cdm = cdm,
  denominatorTable = "denominator",
  outcomeTable = "outcome",
  interval = "years",
  outcomeWashout = Inf,
  repeatedEvents = FALSE
)

# Extract tidy rate table
rates_df <- visOmopResults::tidy(inc) |>
  filter(!is.na(incidence_100000_pys)) |>
  mutate(
    year = as.numeric(substr(incidence_start_date, 1, 4)),
    rate = as.numeric(incidence_100000_pys)
  ) |>
  arrange(year)
```

## Step 4: Fit Segmented Regression (ITS Model)

We define a policy intervention threshold in the calendar year **2000** and construct the segmented time series variables:

```{r fit-its-model, message=FALSE, warning=FALSE}
# Set policy intervention year
intervention_year <- 2000

# Prepare segmented regression design matrix
its_df <- rates_df |>
  mutate(
    time = row_number(),
    post_int = ifelse(year >= intervention_year, 1, 0),
    time_after = ifelse(year >= intervention_year, year - intervention_year, 0)
  )

# Fit Segmented Linear Regression
its_model <- lm(rate ~ time + post_int + time_after, data = its_df)
its_df$fitted <- predict(its_model)

# Format regression coefficients
coefs <- summary(its_model)$coefficients
res_df <- data.frame(
  Parameter = c(
    "Baseline Intercept (Beta 0)",
    "Pre-intervention Slope (Beta 1)",
    "Immediate Step Change (Beta 2)",
    "Slope Change Post-intervention (Beta 3)"
  ),
  Estimate = round(coefs[, 1], 2),
  Std_Error = round(coefs[, 2], 2),
  t_value = round(coefs[, 3], 2),
  p_value = round(coefs[, 4], 4)
)

res_df |>
  gt() |>
  tab_header(
    title = "Segmented Regression Model Results",
    subtitle = "Impact Evaluation of Regulatory Warning (Year 2000)"
  ) |>
  cols_label(
    Parameter = "Model Parameter",
    Estimate = "Estimate",
    Std_Error = "Std. Error",
    t_value = "t-statistic",
    p_value = "p-value"
  )
```

## Step 5: Plot Interrupted Time Series Trends

We plot the empirical annual rate estimates along with the segmented regression trajectories before and after the intervention:

```{r plot-its, message=FALSE, warning=FALSE}
ggplot(its_df, aes(x = year)) +
  geom_point(aes(y = rate), color = "#2b6cb0", size = 2.5) +
  geom_line(aes(y = fitted, group = post_int), color = "#e53e3e", linewidth = 1.2) +
  geom_vline(xintercept = intervention_year - 0.5, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  annotate(
    "text",
    x = intervention_year - 1,
    y = max(its_df$rate) * 0.95,
    label = "Policy Action (2000)",
    hjust = 1,
    fontface = "bold",
    color = "grey30"
  ) +
  labs(
    title = "Interrupted Time Series: Impact Evaluation Analysis",
    subtitle = "Segmented regression of annual GI hemorrhage incidence before vs after regulatory action (2000)",
    x = "Calendar Year",
    y = "Incidence Rate (per 100,000 Person-Years)"
  ) +
  theme_minimal()
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
