---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

Patient-Level Prediction (PLP) aims to develop and validate prognostic models that estimate an individual patient's probability of experiencing a future clinical outcome within a defined time-at-risk (TAR) window based on baseline health history.

## How Patient-Level Prediction Works

- **Target Cohort ($T$)**: Patients at the moment of prediction (e.g. at treatment initiation).
- **Outcome Cohort ($O$)**: The incident binary event to predict within the Time-at-Risk (e.g. Day 1 to 365 post-index).
- **Feature Engineering**: Demographics and historical diagnoses extracted using `PatientProfiles`.
- **Validation**: Independent train/test splitting (`rsample`), model training, discrimination assessment (ROC AUC via `yardstick`), and calibration evaluation.

## Step 1: Setup & Connect to GiBleed

Load the necessary libraries and connect to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(PatientProfiles)
library(rsample)
library(yardstick)
library(ggplot2)
library(gt)
library(dplyr)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Define Target and Outcome Cohorts

We define the target population as new users of NSAIDs (**Celecoxib** `1118084` or **Diclofenac** `1124300`) with at least 365 days of prior observation, and the prediction outcome as incident **Gastrointestinal Hemorrhage** (`192671`):

```{r build-cohorts, message=FALSE, warning=FALSE}
# 1. Target Cohort (T): Incident NSAID users
cdm$target <- conceptCohort(
  cdm = cdm,
  conceptSet = list(nsaid_users = c(1118084L, 1124300L)),
  name = "target"
) |>
  requireIsFirstEntry() |>
  requirePriorObservation(minPriorObservation = 365)

# 2. Outcome Cohort (O): Gastrointestinal Hemorrhage
cdm$outcome <- conceptCohort(
  cdm = cdm,
  conceptSet = list(gi_bleed = 192671L),
  name = "outcome"
)
```

## Step 3: Feature Extraction & Outcome Flagging

We extract demographic variables, baseline comorbidities in the 365-day lookback window ($[-365, 0]$ days), and label whether the outcome occurred during the Time-at-Risk ($[+1, +365]$ days post-index):

```{r extract-features, message=FALSE, warning=FALSE}
# Extract baseline predictors and outcome status in Time-at-Risk (1-365 days)
features_df <- cdm$target |>
  addAge() |>
  addSex() |>
  addPriorObservation() |>
  addConceptIntersectFlag(
    conceptSet = list(
      sinusitis = 4283893L,
      uti = 4116491L,
      asthma = 4051466L
    ),
    window = c(-365, 0)
  ) |>
  addCohortIntersectFlag(
    targetCohortTable = "outcome",
    window = c(1, 365),
    nameStyle = "outcome"
  ) |>
  collect()
```

## Step 4: Model Training and Testing

We partition the dataset into a 75% training set and a 25% testing set using stratified sampling, train a logistic regression risk model, and evaluate test-set performance:

```{r train-and-evaluate, message=FALSE, warning=FALSE}
# Prepare data for modeling
set.seed(42)
features_df$outcome_fac <- factor(
  ifelse(features_df$outcome == 1, "Event", "NoEvent"),
  levels = c("Event", "NoEvent")
)

# 75/25 Train-Test Split
split <- initial_split(features_df, prop = 0.75, strata = outcome_fac)
train_df <- training(split)
test_df <- testing(split)

# Fit prognostic risk model
model <- glm(
  outcome ~ age + sex + prior_observation + sinusitis_m365_to_0 + uti_m365_to_0 + asthma_m365_to_0,
  data = train_df,
  family = binomial()
)

# Generate test set predictions
test_df$pred_prob <- predict(model, newdata = test_df, type = "response")
test_df$pred_class <- factor(
  ifelse(test_df$pred_prob >= 0.18, "Event", "NoEvent"),
  levels = c("Event", "NoEvent")
)

# Compute ROC AUC
auc_val <- roc_auc(test_df, truth = outcome_fac, pred_prob, event_level = "first")
```

## Step 5: Discrimination ROC Curve & Performance Summary

We plot the Receiver Operating Characteristic (ROC) curve and generate a publication-ready performance table:

```{r plot-roc, message=FALSE, warning=FALSE}
# Generate ROC Curve
roc_df <- roc_curve(test_df, truth = outcome_fac, pred_prob, event_level = "first")

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(color = "#2b6cb0", linewidth = 1.2) +
  geom_abline(lty = 3, color = "grey50") +
  coord_equal() +
  labs(
    title = "Patient-Level Prediction: ROC Curve",
    subtitle = paste0("Model: Logistic Regression | Discrimination AUC = ", round(auc_val$.estimate, 3)),
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal()
```

```{r performance-table, message=FALSE, warning=FALSE}
# Performance metric summary table
perf_df <- data.frame(
  Metric = c("ROC AUC (Discrimination)", "Accuracy", "Sensitivity", "Specificity"),
  Estimate = c(
    round(auc_val$.estimate, 3),
    round(accuracy(test_df, truth = outcome_fac, estimate = pred_class)$.estimate, 3),
    round(sens(test_df, truth = outcome_fac, estimate = pred_class, event_level = "first")$.estimate, 3),
    round(spec(test_df, truth = outcome_fac, estimate = pred_class, event_level = "first")$.estimate, 3)
  )
)

perf_df |>
  gt() |>
  tab_header(
    title = "Model Performance Metrics",
    subtitle = "Internal Validation on 25% Holdout Test Set"
  ) |>
  cols_label(
    Metric = "Performance Metric",
    Estimate = "Test Estimate"
  )
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
