---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

Let's implement the study using the OHDSI CohortCharacteristics and PatientProfiles packages to generate baseline profiles (“Table 1”) of one or more cohorts in the OMOP CDM, directly addressing the guide’s descriptive design.

## How CohortCharacteristics works

- **Summarise -> table/plot**: `summariseCohortCount`/`summariseCharacteristics` produce tidy `summarised_result` outputs consumed by `table…`/`plot…` helpers in `visOmopResults`.
- **Cohorts in, summaries out**: You provide one or more cohort tables (OMOP CDM structure). Functions compute counts, attrition, demographics, and clinical features at/before index.
- **Windows**: For granular baselines, large-scale summaries can use windows (e.g., -365:-1, 0:0, 1:365) around index.
- **Groups/strata**: Each cohort produces groups; you can stratify or facet by `cohort_name` and other variables in plots/tables.

## Practical guidance

- Connect to standard benchmark datasets (such as Eunomia GiBleed) for reproducible evaluation.
- Keep cohort inclusion/exclusion logic in cohort construction; keep `summarise` functions descriptive.
- Save the `summarised_result` object and then render tables/plots.

## Step 1: Setup

This initial step focuses on preparing the R environment for the analysis. It involves loading libraries like `CDMConnector`, `CohortConstructor`, `CohortCharacteristics`, and `PatientProfiles`.

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(CohortCharacteristics)
library(PatientProfiles)
library(visOmopResults)
library(dplyr)
```

## Step 2: Connect to the GiBleed CDM and Create Cohorts

We connect to the standardized Eunomia **GiBleed** dataset using DuckDB. We define two comparative cohorts of interest—individuals exposed to **Celecoxib** (`concept_id = 1118084`) and **Diclofenac** (`concept_id = 1124300`).

```{r connect-gibleed-and-create-cohorts, message=FALSE, warning=FALSE}
# Ensure GiBleed dataset is available and connect via DuckDB
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")

# Generate cohorts for Celecoxib and Diclofenac
cdm$nsaids <- conceptCohort(
  cdm = cdm,
  conceptSet = list(
    celecoxib = 1118084L,
    diclofenac = 1124300L
  ),
  name = "nsaids"
)
```

## Step 3: Cohort counts and attrition

Before diving into detailed characterization, it's important to understand the size and composition of each cohort. This step calculates the number of subjects and records in each cohort and inspects the attrition process.

```{r summarize-cohort-counts, message=FALSE, warning=FALSE}
# Summarize the number of subjects and records in each cohort
counts <- summariseCohortCount(cdm$nsaids)
# Display the cohort counts in a table
tableCohortCount(counts)
```

```{r summarize-cohort-attrition, message=FALSE, warning=FALSE}
# Summarize the attrition of the cohorts
attr <- summariseCohortAttrition(cdm$nsaids)
# Display the cohort attrition in a table
tableCohortAttrition(attr)
```

## Step 4: Baseline characteristics (“Table 1”) and plots

This is the core of the patient-level characterization analysis. We generate a comprehensive summary of baseline characteristics, often referred to as "Table 1," describing age, sex, and prior observation time at cohort entry.

```{r summarize-baseline-characteristics, message=FALSE, warning=FALSE}
# Summarize baseline characteristics (demographics)
chars <- summariseCharacteristics(cdm$nsaids)
# Display baseline characteristics table
tableCharacteristics(chars)
```

```{r plot-age-distribution, message=FALSE, warning=FALSE}
# Create a comparative boxplot of the age distribution across cohorts
plot_data <- dplyr::filter(chars, variable_name == "Age")
plotCharacteristics(plot_data, plotType = "boxplot", colour = "cohort_name")
```

## Step 5: Advanced Characterization with Comorbidity Flags

The `summariseCharacteristics` function allows creating custom clinical flags with `conceptIntersectFlag` to assess pre-existing conditions in the baseline lookback window ($[-365, 0]$ days prior to cohort entry).

```{r summarize-characteristics-with-flags, message=FALSE, warning=FALSE}
# Summarize baseline comorbidity flags in the 365 days prior to index
flags_summary <- cdm$nsaids |>
  summariseCharacteristics(
    demographics = TRUE,
    conceptIntersectFlag = list(
      "Baseline Comorbidities" = list(
        conceptSet = list(
          sinusitis = 4283893L,
          gi_bleed = 192671L,
          uti = 4116491L,
          asthma = 4051466L
        ),
        window = c(-365, 0)
      )
    )
  )

# Display the enriched characteristics table
tableCharacteristics(flags_summary)
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```