---
knit: |
  (function(inputFile, encoding) {
    # Render the document to markdown
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output',
      paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

The OHDSI [`IncidencePrevalence`](/docs/data_analysis/package_reference/IncidencePrevalence) R package is designed to calculate population-level incidence and prevalence rates from OMOP CDM data, directly addressing the core analytical requirements of this study design.

At a high level, IncidencePrevalence turns OMOP CDM cohorts into tidy, stratified summaries by repeatedly applying three ideas:

- **Denominator time:** People contribute at-risk time only when they meet the study rules you set with `generateDenominatorCohortSet()`/`generateTargetDenominatorCohortSet()`. Entry is the latest of study start, prior observation met, and minimum age; exit is the earliest of study end, observation end, and age upper limit.
- **Outcome overlap:** Outcomes are read from a separate cohort table you prepare (here provided by `mockIncidencePrevalence()`). Prevalence checks whether a person is in the outcome cohort at a time point/over an interval; incidence counts first qualifying onsets, respecting washout and repeated event rules.
- **Interval slicing:** The study time is cut into calendar intervals (years, quarters, months). Each interval gets its own denominator, outcome counts, and derived measures.

### Key computations and arguments

- **Point prevalence (`estimatePointPrevalence`):**
  - **What:** proportion with the outcome at a time point inside each interval.
  - **How:** outcome presence on the chosen `timePoint` (“start”/“middle”/“end”) among those in the denominator at that date.
  - **Tip:** Use when you want a snapshot at a consistent anchor within each interval.

- **Period prevalence (`estimatePeriodPrevalence`):**
  - **What:** proportion with the outcome at any time during the interval.
  - **How:** outcome occurs at any day overlapping the interval; denominator includes those present during the interval.
  - `fullContribution`: if `TRUE`, only people observed for the entire interval count in the denominator; if `FALSE`, anyone observed for ≥1 day counts.
  - **Tip:** Choose `fullContribution` based on whether incomplete observation could bias proportions for long intervals.

- **Incidence (`estimateIncidence`):**
  - **What:** new-onset rates per person-time in each interval.
  - **How:** counts qualifying outcome onsets and divides by person-time contributed in the denominator.
  - `outcomeWashout`: requires no outcome in the prior N days; use `Inf` to ensure first-ever incidence.
  - `repeatedEvents`: if `TRUE`, allows multiple incident events per person separated by washout; if `FALSE`, at most one incident event per person.
  - `completeDatabaseIntervals`: if `TRUE`, drops intervals where the database cannot establish full observation coverage (e.g., first/last years with partial data), reducing edge bias.

### Stratification and groups

- **Groups:** Each combination of denominator settings (age group, sex, prior history, time-at-risk windows, target-based subsets) becomes a separate `cohort_definition_id` and produces separate rows.
- **Strata:** You can facet/colour or summarize by variables such as `denominator_age_group`, `denominator_sex`, etc. Use `plotIncidence()`/`plotPrevalence()` helpers and `available...Grouping()` to see valid aesthetics.

### Outputs

- **Summarised results:** Functions return a “summarised_result” tables with `estimate_name`/`value` pairs, plus additional metadata (start/end dates, interval type). These are directly consumable by `visOmopResults` to `tableIncidence`/`tablePrevalence` and plot functions.

## Practical guidance

- Pick intervals aligned to your question and data density (years for long-term trends, months for seasonality).
- Use `outcomeWashout` and `repeatedEvents` to encode “first-ever” vs. “episode-based” incidence.
- Prefer `completeDatabaseIntervals` (incidence) and consider `fullContribution` (period prevalence) when edge effects or intermittent observation could bias estimates.
- Keep exclusion logic out of outcome cohorts; define restrictions on the denominator instead.

Our workflow will follow a simple, powerful pattern common across OHDSI packages: **`summarise -> table/plot`**.

### Step 1: Load Libraries & Connect to the GiBleed Database

First, we load the necessary libraries. `CDMConnector` provides the connection to the OMOP CDM database, `CohortConstructor` builds clinical cohorts from standard OMOP concepts, `IncidencePrevalence` calculates population-level epidemiology rates, and `visOmopResults` formats the results into publication-ready tables and plots.

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(  
  fig.path = "../../assets/images/rmd_output/"  
)
```

We connect to the standardized Eunomia **GiBleed** dataset using DuckDB:

```{r load_libraries}
library(CDMConnector)
library(CohortConstructor)
library(IncidencePrevalence)
library(visOmopResults)
library(dplyr)
```

```{r connect_gibleed, message=FALSE, warning=FALSE}
# Ensure GiBleed dataset is available and connect via DuckDB
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

### Step 2: Define Outcome and Denominator Cohorts

In this study, our outcome of interest is **Gastrointestinal Hemorrhage** (`concept_id = 192671`). We instantiate the outcome cohort directly from the OMOP CDM condition table:

```{r define_outcome_cohort, message=FALSE, warning=FALSE}
# Create the gastrointestinal hemorrhage outcome cohort
cdm$outcome <- conceptCohort(
  cdm = cdm,
  conceptSet = list(gi_bleed = 192671L),
  name = "outcome"
)
```

Next, we define our source study population (the denominator). We will study the population observed between 1980 and 2019, stratifying by age groups (0–49 and 50–100) and by sex:

```{r define_denominator_cohort, message=FALSE, warning=FALSE}
# Define the denominator cohort for the study period
cdm <- generateDenominatorCohortSet(
  cdm = cdm,
  name = "denominator",
  cohortDateRange = as.Date(c("1980-01-01", "2019-01-01")),
  ageGroup = list(
    c(0, 49),
    c(50, 100)
  ),
  sex = c("Male", "Female")
)
```

### Step 3: Calculate Incidence & Prevalence (The `summarise` Step)

This step calculates incidence and prevalence rates across the defined strata.

#### 3.1: Estimate Incidence

We calculate annual incidence rates for gastrointestinal hemorrhage. We set `outcomeWashout = Inf` to include only first-ever occurrences and `repeatedEvents = FALSE`. We then visualize the results with `plotIncidence()` and generate a structured table with `tableIncidence()`:

```{r calculate_incidence_and_plot, echo=TRUE, message=FALSE, warning=FALSE}
# Calculate annual incidence rates
inc <- estimateIncidence(
  cdm = cdm,
  denominatorTable = "denominator",
  outcomeTable = "outcome",
  interval = "years",
  outcomeWashout = Inf,
  repeatedEvents = FALSE
)

# Plot incidence rates faceted by age group and colored by sex
plotIncidence(inc, facet = "denominator_age_group", colour = "denominator_sex")
```

```{r table_incidence, echo=TRUE, message=FALSE, warning=FALSE}
# Display incidence summary table
tableIncidence(inc)
```

#### 3.2: Estimate Point Prevalence

Next, we estimate annual point prevalence at the start of each year (`timePoint = "start"`):

```{r calculate_prevalence_and_plot, echo=TRUE, message=FALSE, warning=FALSE}
# Calculate annual point prevalence
prev_point <- estimatePointPrevalence(
  cdm = cdm,
  denominatorTable = "denominator",
  outcomeTable = "outcome",
  interval = "years",
  timePoint = "start"
)

# Plot point prevalence faceted by age group and colored by sex
plotPrevalence(prev_point, facet = "denominator_age_group", colour = "denominator_sex")
```

```{r table_prevalence, echo=TRUE, message=FALSE, warning=FALSE}
# Display prevalence summary table
tablePrevalence(prev_point)
```

```{r disconnect_db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```