---
knit: |
  (function(inputFile, encoding) {
    # Render the document to markdown
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output',
      paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

```{r setup-knitr, include=FALSE}
# Set up knitr options for knitting the Rmd file.
# The upload.fun option modifies the image path for embedding in the final document.
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
# The fig.path option sets the directory where plots generated by the R code chunks will be saved.
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

# R Programming Basics

This guide introduces the fundamentals of R programming, focusing on data manipulation with the **tidyverse**, particularly `dplyr` and `dbplyr`. These basics are essential before diving into OMOP-specific analyses.

## Who is this guide for?

This guide is written for anyone who wants to work with databases in a “tidyverse” style—a human-centered, consistent, and composable approach to data analysis.

- **New to R?** We recommend complementing this guide with [*R for Data Science*](https://r4ds.had.co.nz/).
- **New to databases?** Familiarize yourself with SQL basics through tutorials like [SQLBolt](https://sqlbolt.com/) or [SQLZoo](https://www.sqlzoo.net/wiki/SQL_Tutorial).
- **New to the OMOP CDM?** This guide is best paired with [*The Book of OHDSI*](https://ohdsi.github.io/TheBookOfOhdsi/).

## Bridging the Gap for Different Backgrounds

This guide is designed to be accessible to readers from various backgrounds. Here’s how we address common challenges:

### For SAS Programmers

If you’re coming from SAS, you might find the functional programming style of R unfamiliar. Here’s a quick mapping of common SAS concepts to their R/dplyr equivalents:

| SAS Concept | R/dplyr Equivalent | Description |
|----|----|----|
| `PROC SQL` | `dplyr` verbs like `filter()`, `select()`, `mutate()` | Querying and manipulating data |
| `DATA step` | `mutate()` or `transmute()` | Creating or modifying variables |
| `PROC SORT` | `arrange()` | Sorting data |
| `PROC SUMMARY` | `group_by()` + `summarise()` | Grouping and aggregating data |
| `PROC FREQ` | `count()` | Frequency tables |

Remember, in R, operations are chained using the pipe (`|>`) for readability, similar to how you might chain procedures in SAS.

### For Clinical Experts

If you’re new to data analysis but have a strong clinical background, the OMOP CDM might seem overly complex at first. Think of it as a standardized "language" for health data:

- **Why separate tables?** Instead of one big spreadsheet, data is split into logical tables (e.g., `person` for demographics, `condition_occurrence` for diagnoses) to avoid repetition and ensure consistency.
- **What are concept IDs?** These are standardized codes for medical terms (e.g., "Type 2 Diabetes" might have ID 201826). They ensure everyone uses the same definitions, making research comparable.
- **Why vocabularies?** OMOP uses controlled vocabularies (like SNOMED or ICD) to map local codes to standard ones, reducing ambiguity.

This structure allows for powerful, scalable analyses across different healthcare systems.

### For Data Scientists New to Healthcare

If you’re proficient in R but unfamiliar with clinical research, we’ll explain the “why” behind the code. For example, when we create a “cohort” of patients, it’s not just filtering data—it’s defining a study population based on clinical criteria to answer specific research questions.

## Detailed Chapters Overview

#### **1. A first analysis using data in a database**

The `iris` dataset is a classic example in data science, collected by Edgar Anderson in 1935. It contains measurements of 150 iris flowers from three species: setosa, versicolor, and virginica. Each flower has four measurements: sepal length, sepal width, petal length, and petal width (all in centimeters). The goal is often to classify the species based on these measurements, making it a simple but effective dataset for demonstrating statistical and machine learning techniques.

### Setting Up Your Environment

First, you need to install and load the necessary R packages.

```{r load-libraries-general, message=FALSE, warning=FALSE}
# Load the libraries
library(dplyr)
library(dbplyr)
library(ggplot2)
library(DBI)
library(duckdb)
```

### Inserting Data into a Database

Next, we will load the `iris` data into an in-memory `duckdb` database. This simulates a real-world scenario where your data resides in a database server.

```{r insert-iris}
# Create an in-memory duckdb database
db <- dbConnect(drv = duckdb())

# Write the iris dataframe to a table named "iris" in the database
dbWriteTable(db, "iris", iris)

# You can see the tables in the database
dbListTables(db)
# > [1] "iris"
```

### From R to SQL: The Power of `dbplyr`

Now that the data is in a database, we could query it using SQL. However, the magic of the `dbplyr` package is that it allows you to write familiar `dplyr` code, which it automatically translates into SQL for you.

First, create a reference to the `iris` table in the database.

```{r create-iris-db}
iris_db <- tbl(db, "iris")
```

Now, you can use standard `dplyr` verbs on this object.

```{r dbplyr-example}
# Get a summary of sepal length by species
iris_db |>
  group_by(Species) |>
  summarise(
    n = n(),
    mean_sepal_length = mean(Sepal.Length, na.rm = TRUE)
  )
```

When you run this code, `dbplyr` translates it into the following SQL query, sends it to the database, and returns the result.

```sql
<SQL>
SELECT
  "Species",
  COUNT(*) AS n,
  AVG("Sepal.Length") AS mean_sepal_length
FROM iris
GROUP BY "Species"
```

This allows you to perform complex data manipulations using intuitive R code, without needing to write complex, database-specific SQL.

### Bringing Data into R for Visualization

While most data manipulation should happen in the database for efficiency, you will often need to bring the final, summarized data back into R for visualization or modeling. The `collect()` function does this.

Let's create a histogram of sepal length for each flower species.

```{r visualization}
iris_db |>
  select("Species", "Sepal.Length") |>
  collect() |> # This brings the data from the database into an R dataframe
  ggplot(aes(x = `Sepal.Length`, fill = Species)) +
  geom_histogram(binwidth = 0.2, colour = "black") +
  facet_wrap(~Species, ncol = 1) +
  theme_bw()
```

This workflow—manipulating data in the database and collecting only the results—is the most efficient way to work with large datasets.

### Disconnecting from the Database

When you are finished with your analysis, it is good practice to close the connection to the database.

```{r disconnect}
dbDisconnect(db)
```

#### **2. Core verbs for analytic pipelines utilising a database**

To demonstrate working with multiple tables, we'll extend the `iris` example by creating related tables. We'll split the data into separate tables for species information and measurements, then show how to join them back together.

### Setting Up the Database with Multiple Tables

```{r setup-chapter2}
# Create database
db <- dbConnect(duckdb())

# Split iris into two tables: species info and measurements
species_info <- iris |> select(Species) |> distinct() |> mutate(species_id = row_number())
measurements <- iris |> left_join(species_info, by = "Species")

# Write tables
dbWriteTable(db, "species_info", species_info)
dbWriteTable(db, "measurements", measurements)

# Create lazy references
species_db <- tbl(db, "species_info")
measurements_db <- tbl(db, "measurements")
```

### Selecting Rows: `filter()` and `distinct()`

Use `filter()` to select rows based on conditions. For example, to find measurements with sepal length > 5:

```{r filter-example}
measurements_db |>
  filter(Sepal.Length > 5) |>
  select(Species, Sepal.Length)
```

`distinct()` removes duplicate rows:

```{r distinct-example}
measurements_db |>
  distinct(Species)
```

### Ordering Rows: `arrange()`

`arrange()` sorts rows by one or more columns:

```{r arrange}
measurements_db |>
  arrange(desc(Sepal.Length)) |>
  select(Species, Sepal.Length)
```

### Column Transformation: `mutate()`, `select()`, `rename()`

`mutate()` creates or modifies columns:

```{r mutate-example}
measurements_db |>
  mutate(
    sepal_ratio = Sepal.Length / Sepal.Width,
    is_large = Sepal.Length > 6
  ) |>
  select(Species, sepal_ratio, is_large)
```

`select()` chooses specific columns:

```{r select-example}
measurements_db |>
  select(Species, Sepal.Length, Petal.Length)
```

`rename()` changes column names:

```{r rename-example}
measurements_db |>
  rename(sepal_length = Sepal.Length)
```

### Grouping and Aggregation: `group_by()`, `summarise()`, `count()`

`group_by()` groups data for calculations:

```{r group-summarise-example}
measurements_db |>
  group_by(Species) |>
  summarise(
    avg_sepal = mean(Sepal.Length, na.rm = TRUE),
    count = n()
  )
```

`count()` is a shortcut for counting:

```{r count-example}
measurements_db |>
  count(Species, sort = TRUE)
```

### Joining Tables

Joins combine data from multiple tables. For example, to join measurements with species info:

```{r joins}
measurements_db |>
  left_join(species_db, by = "Species") |>
  select(Species, Sepal.Length)
```

### Constructing a Tidy Analytic Dataset

Let's create a comprehensive dataset by joining tables and performing aggregations:

```{r analytic-dataset}
analytic_dataset <- measurements_db |>
  group_by(Species) |>
  summarise(
    avg_sepal_length = mean(Sepal.Length, na.rm = TRUE),
    avg_petal_length = mean(Petal.Length, na.rm = TRUE),
    total_measurements = n()
  ) |>
  arrange(desc(avg_sepal_length))

# Collect results
analytic_dataset |> collect()
```

This pipeline demonstrates how to build complex queries using `dplyr` verbs, with all operations pushed to the database for efficiency.

#### **3. Supported expressions for database queries**

### Data Type Conversions

R and SQL handle data types differently. `dbplyr` automatically converts many types, but understanding the mappings is important:

- Logical: `TRUE`/`FALSE` → `TRUE`/`FALSE` or `1`/`0`
- Character: Strings remain strings
- Numeric: Integers and doubles are preserved
- Dates: `Date` objects → `DATE`
- Datetimes: `POSIXct` → `TIMESTAMP`

### Logical Comparisons and Operators

Standard comparison operators work as expected:

```{r logical-ops}
measurements_db |> filter(Sepal.Length > 5 & Petal.Length > 3)
# Translates to: WHERE "Sepal.Length" > 5 AND "Petal.Length" > 3
```

### Conditional Statements

`if_else()` and `case_when()` are supported:

```{r conditional}
measurements_db |>
  mutate(
    size_category = case_when(
      Sepal.Length < 5 ~ "small",
      Sepal.Length <= 6 ~ "medium",
      TRUE ~ "large"
    )
  )
```

SQL translation:

```sql
CASE
  WHEN "Sepal.Length" < 5 THEN 'small'
  WHEN "Sepal.Length" <= 6 THEN 'medium'
  ELSE 'large'
END
```

### String Functions

Common string operations:

```{r string-funcs}
species_db |>
  mutate(
    species_upper = toupper(Species),
    species_length = nchar(Species),
    species_substr = substr(Species, 1, 3)
  )
```

### Database-Specific Considerations

While `dbplyr` aims for portability, some functions may behave differently:

- DuckDB: Full support for most R functions
- PostgreSQL: Excellent date/time support
- SQL Server: May require workarounds for some functions

Always test your queries on your target database system.

#### **4. Building analytic pipelines for a data model**

### Principles of Modular Pipelines

1. **Separation of Concerns**: Break down complex queries into smaller, focused steps
2. **Reusability**: Create functions for common operations
3. **Readability**: Use clear variable names and comments
4. **Efficiency**: Minimize data movement between database and R

### Example: Analyzing Iris Measurements

Building on the `iris` example, let's create a pipeline to analyze measurements by species:

```{r pipeline-example}
# Step 1: Filter and prepare data
iris_performance <- measurements_db |>
  filter(!is.na(Sepal.Length), !is.na(Petal.Length))

# Step 2: Calculate metrics
performance_metrics <- iris_performance |>
  mutate(
    sepal_petal_ratio = Sepal.Length / Petal.Length,
    is_large_flower = as.numeric(Sepal.Length > 6)
  ) |>
  group_by(Species) |>
  summarise(
    total_flowers = n(),
    avg_sepal_petal_ratio = mean(sepal_petal_ratio, na.rm = TRUE),
    large_percentage = mean(is_large_flower, na.rm = TRUE) * 100
  )

# Step 3: Filter for significant groups
significant_species <- performance_metrics |>
  filter(total_flowers >= 40) |>
  arrange(desc(avg_sepal_petal_ratio))

# Collect results
results <- significant_species |> collect()
```

### Creating Reusable Functions

For repeated operations, create functions:

```{r reusable-functions}
calculate_ratios <- function(measurements_tbl) {
  measurements_tbl |>
    mutate(
      sepal_petal_ratio = Sepal.Length / Petal.Length,
      is_large = as.numeric(Sepal.Length > 6)
    ) |>
    summarise(
      total = n(),
      avg_ratio = mean(sepal_petal_ratio, na.rm = TRUE),
      large_pct = mean(is_large, na.rm = TRUE)
    )
}

# Use the function
species_ratios <- measurements_db |>
  group_by(Species) |>
  calculate_ratios()
```

### Handling Complex Joins in Relational Models

In structured data models like OMOP, you'll often need to join multiple tables. Plan your joins carefully:

1. Identify the central table (e.g., `person` in OMOP)
2. Determine the join keys
3. Consider the join type (left, inner, etc.)
4. Chain joins logically

This approach ensures your pipelines are maintainable and efficient.
