---
knit: |
  (function(inputFile, encoding) {
    rmarkdown::render(
      inputFile,
      encoding = encoding,
      output_format = rmarkdown::github_document(html_preview = FALSE),
      output_file = file.path('../../_includes/rmd_output', paste0(tools::file_path_sans_ext(basename(inputFile)), '.md'))
    )
  })
---

Pathway Analysis maps out the real-world sequence, combination, and transition of therapies or clinical events experienced by patients over time following an initial disease diagnosis.

## How Pathway Analysis Works

- **Target Cohort**: Patients diagnosed with a condition of interest establishing index "time zero".
- **Event Extraction & Era Building**: Extracting subsequent drug exposures, collapsing adjacent records into treatment eras.
- **Line of Therapy Assignment**: Chronologically ordering treatment eras into Line 1 (L1), Line 2 (L2), and Line 3 (L3).
- **Pathway Aggregation & Visualisation**: Quantifying the most frequent therapeutic sequences and visualizing patient journey flows.

## Step 1: Setup & Connect to GiBleed

Load the necessary libraries and establish a connection to the Eunomia **GiBleed** dataset using DuckDB:

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(upload.fun = function(x) sub("^\\.\\./\\.\\./", "/", x))
knitr::opts_chunk$set(fig.path = "../../assets/images/rmd_output/")
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
```

```{r connect-gibleed, message=FALSE, warning=FALSE}
# Connect to Eunomia GiBleed dataset
Sys.setenv(EUNOMIA_DATA_FOLDER = Sys.getenv("EUNOMIA_DATA_FOLDER", tempdir()))
if (!eunomiaIsAvailable("GiBleed")) {
  downloadEunomiaData("GiBleed")
}

con <- DBI::dbConnect(duckdb::duckdb(), eunomiaDir("GiBleed"))
cdm <- cdmFromCon(con, cdmSchema = "main", writeSchema = "main")
```

## Step 2: Define Index Diagnosis Cohort

We define our target cohort as incident patients diagnosed with **Sinusitis** (`concept_id = 4283893, 4294548`):

```{r define-target-cohort, message=FALSE, warning=FALSE}
# Target Cohort: Incident Sinusitis Patients
cdm$sinusitis <- conceptCohort(
  cdm = cdm,
  conceptSet = list(sinusitis = c(4283893L, 4294548L)),
  name = "sinusitis"
) |>
  requireIsFirstEntry()
```

## Step 3: Extract Longitudinal Medications & Construct Lines of Therapy

We extract all subsequent drug exposures following the diagnosis date, consolidate consecutive exposures into treatment eras, and classify each patient's journey into sequential lines of therapy:

```{r construct-pathways, message=FALSE, warning=FALSE}
# Extract post-diagnosis drug exposures
drugs <- cdm$sinusitis |>
  select(person_id = subject_id, diag_date = cohort_start_date) |>
  inner_join(
    cdm$drug_exposure |> select(person_id, drug_concept_id, drug_exposure_start_date),
    by = "person_id"
  ) |>
  filter(drug_exposure_start_date >= diag_date) |>
  left_join(cdm$concept |> select(concept_id, concept_name), by = c("drug_concept_id" = "concept_id")) |>
  collect() |>
  mutate(
    treatment = case_when(
      grepl("Ampicillin", concept_name, ignore.case = TRUE) ~ "Ampicillin",
      grepl("celecoxib", concept_name, ignore.case = TRUE) ~ "Celecoxib",
      grepl("Diclofenac", concept_name, ignore.case = TRUE) ~ "Diclofenac",
      grepl("vaccine|toxoid", concept_name, ignore.case = TRUE) ~ "Vaccines",
      TRUE ~ "Other"
    )
  ) |>
  arrange(person_id, drug_exposure_start_date) |>
  group_by(person_id) |>
  mutate(prev_treatment = lag(treatment)) |>
  filter(is.na(prev_treatment) | treatment != prev_treatment) |>
  mutate(line_num = row_number()) |>
  filter(line_num <= 3) |>
  mutate(line = paste0("Line_", line_num)) |>
  ungroup()

# Pivot to patient-level pathways
pathways <- drugs |>
  pivot_wider(id_cols = person_id, names_from = line, values_from = treatment) |>
  mutate(
    Line_1 = ifelse(is.na(Line_1), "None", Line_1),
    Line_2 = ifelse(is.na(Line_2), "Discontinued", Line_2),
    Line_3 = ifelse(is.na(Line_3), "Discontinued", Line_3)
  )
```

## Step 4: Top Treatment Pathways Summary Table

We aggregate the individual patient journeys into top treatment pathways:

```{r pathway-table, message=FALSE, warning=FALSE}
# Summarise most common treatment pathways
pathway_summary <- pathways |>
  group_by(Line_1, Line_2, Line_3) |>
  summarise(Patients_N = n(), .groups = "drop") |>
  mutate(Percentage = round((Patients_N / sum(Patients_N)) * 100, 1)) |>
  arrange(desc(Patients_N)) |>
  head(10)

pathway_summary |>
  gt() |>
  tab_header(
    title = "Top Treatment Pathways Following Sinusitis Diagnosis",
    subtitle = "Sequence Progression across Lines 1, 2, and 3"
  ) |>
  cols_label(
    Line_1 = "1st Line Therapy",
    Line_2 = "2nd Line Therapy",
    Line_3 = "3rd Line Therapy",
    Patients_N = "Patients (N)",
    Percentage = "Proportion (%)"
  )
```

## Step 5: Visualizing Treatment Flow across Lines

We plot the distribution of therapeutic modalities across successive lines of therapy:

```{r plot-pathway, message=FALSE, warning=FALSE}
# Aggregate by line and treatment
plot_df <- drugs |>
  group_by(line, treatment) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(line) |>
  mutate(pct = (n / sum(n)) * 100)

ggplot(plot_df, aes(x = line, y = pct, fill = treatment)) +
  geom_col(position = "stack", width = 0.6, color = "white") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Treatment Pathway Distribution Across Lines of Therapy",
    subtitle = "Longitudinal pharmacotherapy progression following initial sinusitis diagnosis",
    x = "Line of Therapy",
    y = "Proportion of Patients (%)",
    fill = "Treatment"
  ) +
  theme_minimal()
```

```{r disconnect-db, include=FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
